Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-26 NovelLBM.tex
upload log files

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication but can also serve as a common food additive.
Our solution contains \SI{4}{\percent} alginate dissolved in PBS and was measured in a commercial plate-plate rheometer (see appendix~\ref{app:experiment} for experimental details).
The shear-dependent viscosity $\eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than \SI{10000}{\milli\pascal\second}.
Bioinks in biofabrication are in general characterized by such high viscosities to guarantee stability of the printed construct.
At the same time, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
This motivates the use of highly shear-thinning viscoelastic fluids in biofabrication.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_N.pdf}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model. }
		\label{fig:fit_alginate}
	\end{figure}

The second cell carrier fluid modeled in this work is a methyl cellulose solution which is commonly used in real-time-deformation-cytometry (RT-DC) experiments.
RT-DC experiments allow very high-throughput measurements of cell mechanical properties and have been used successfully for disease detection by screening blood samples.
We extract our data from \cite{buyukurganci2023} where the shear rheology of methyl cellulose was measured using a plate-plate and cone-plate rheometer.
Three concentrations of \SI{0.49}{\percent}, \SI{0.59}{\percent} and \SI{0.83}{\percent} methyl cellulose in PBS were investigated, and their rheology is shown in figure~\ref{fig:fit_mc}.
Even at the highest concentration the zero-shear viscosity is two orders of magnitude lower than for the considered alginate solution.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_mc_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_mc_N.pdf}
		\caption{The viscosity (a) and normal stress (b) of a methyl cellulose solution from \cite{buyukurganci2023}.
				Lines show a fit of the PTT rheological model. }
		\label{fig:fit_mc}
	\end{figure}

@@

The PTT model is specified by a differential equation for the polymer stress $\ma{\tau}$:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
We note that extended forms of the PTT model with a Mittag-Leffler instead of an exponential function have been studied as well \cite{ferras2019}.
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at what shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \vec{u}\cdot\nabla\ma{A}.
	\end{equation}
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function. \SG{ check that $ \eta _ \mathrm{s}$ is correct }\RK{what do you mean by correct?} 
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_ \eta $\\
			\hline
			alginate \num{4}\% & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & $4.8 \times 10 ^{ 4 }$\\
			methyl c. \num{0.49}\% & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & 18.7\\
			methyl c. \num{0.59}\% & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & 32.5\\
			methyl c. \num{0.83}\% & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & 81
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
		\SG{ add a column with $R_ \eta $ }}\RK{would be identical to the content of the $\eta_p$ column, also would be used before introduced} \SG{ Stimmt, aber ich will es auf einen Blick sichtbar machen }
		\label{tab:PTTparameters}
	\end{table}

We introduce the ratio between the polymer and the solvent viscosity as
\begin{align}
	R_\eta = \frac{\eta_\text{p}}{\eta_\text{s}}		\label{eq:R_eta}
\end{align}
which will turn out to be an important parameter to determine numerical stability of LB simulations.
\SG{ wo don't actually show that $R_ \eta  $ determines stability? }\RK{we do not. We could do a few simulations with $\alpha_s = 0$ and increasing $R_\eta$ until the sim breaks if you want} \SG{ wie besprochen, bau einfach die Zitate ein, das reicht dann. }
This ratio can similarly be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979}.
For our alginate solution we obtain $R_\eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological data from various literature sources have been compiled into a single format to determine $R_\eta$.
The \SI{0.49}{\percent} methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.

\begin{table}[H]
	\centering
		\begin{tabular}{l|p{3cm}|p{1.5cm}|p{1.5cm}|p{1.5cm}|p{2cm}}
			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  >1e4 & <3e-1 & >3e4 & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & 3e2 & 1e-4 & 3e6 & Table 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& 5e4 & <5e1 & >1e3 & Fig. 4b \\
			\rowcolor{secondaryBackground}
			O'Connell\cite{oconnell2020} & pluronic F-127 & >1e6 & <8 & > 2e5 & Chapter 7 Fig. 13 \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids.
	\SG{ cross-check these values }\RK{as in with other literature?}
	}
	\label{tab:bioinks}
\end{table}
\RK{While $\eta_0/\eta_\infty$ would have been sensible for $R_\eta$ definition, i used $(\eta_0-\eta_\infty)/\eta_\infty$ instead. This plays nicer with our models and does not make a difference, but the notation above is now inconsistent since del of CY}

@@

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }\RK{dellar is missing}
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }\RK{gupta2015\_2 not in intro} \SG{ dann setz einfach die Zitate so rein, wie du sie fuer passend haeltst. }
Since the early days of LBM, variants for viscoelastic fluids have been developed.
Keeping in mind our target geometries, we focus here on three-dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a basic viscoelastic fluid obeying Jeffery's model \cite{lallemand2003}.
More advanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells.
The work of \cite{wang2020} solved the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM.
The same technique was used by \cite{malaspinas2010} and \cite{ma2020} for the Oldroyd-B and FENE-P models.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \cite{malaspinas2010,ma2020,wang2020,wang2021}, which introduces a small error compared to the analytical solution of the constitutive equation.
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015} \SG{ rephrase and give some more details }\RK{I have not the slightest clue what gupta is doing. I assume optimize is code for increase until stable but I do not know.}.
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
A recent comparison in 2D found that the former in general leads to better stability \cite{dzanic2022_2}.\RK{No in fact, it uses the latter and found to be more accurate than a paper using the former, stability was entirely attributed to the decomposition and artificial diffusivity.} \SG{ die Referenz war falsch, es war der andere Dzanic. Aber da steht eigentlich schon, dass forcing besser sei, oder? }

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose. 
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}

@@

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. \SG{ comment whether these definitions are identical or different than the definitions above. }\RK{different but can be transformed into each other $\frac{1}{R_\nu}-1=R_\eta$}
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should 8 instead of 4. \SG{ stimmt das so? }

>>>>

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication but can also serve as a common food additive.
Our solution contains \SI{4}{\percent} alginate dissolved in PBS and was measured in a commercial plate-plate rheometer (see appendix~\ref{app:experiment} for experimental details).
The shear-dependent viscosity $\eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than \SI{10000}{\milli\pascal\second}.
Bioinks in biofabrication are in general characterized by such high viscosities to guarantee stability of the printed construct.
At the same time, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
This motivates the use of highly shear-thinning viscoelastic fluids in biofabrication.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_N.pdf}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model.}
		\label{fig:fit_alginate}
	\end{figure}

The second cell carrier fluid modeled in this work is a methyl cellulose solution which is commonly used in real-time-deformation-cytometry (RT-DC) experiments.
RT-DC experiments allow very high-throughput measurements of cell mechanical properties and have been used successfully for disease detection by screening blood samples.
We extract our data from \cite{buyukurganci2023} where the shear rheology of methyl cellulose was measured using a plate-plate and cone-plate rheometer.
Three concentrations of \SI{0.49}{\percent}, \SI{0.59}{\percent} and \SI{0.83}{\percent} methyl cellulose in PBS were investigated, and their rheology is shown in figure~\ref{fig:fit_mc}.
Even at the highest concentration the zero-shear viscosity is two orders of magnitude lower than for the considered alginate solution.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_mc_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_mc_N.pdf}
		\caption{The viscosity (a) and normal stress (b) of a methyl cellulose solution from \cite{buyukurganci2023}.
				Lines show a fit of the PTT rheological model.}
		\label{fig:fit_mc}
	\end{figure}

@@

The PTT model is specified by a differential equation for the polymer stress $\ma{\tau}$:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
We note that extended forms of the PTT model with a Mittag-Leffler instead of an exponential function have been studied as well \cite{ferras2019}\todo{not really what we want to say here. We took this formula mostly from there}.
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at what shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \vec{u}\cdot\nabla\ma{A}.
	\end{equation}
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function. \SG{ check that $ \eta _ \mathrm{s}$ is correct }\RK{what do you mean by correct?}
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.8e4} \\
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
		\SG{ add a column with $R_ \eta $ }}\RK{would be identical to the content of the $\eta_p$ column, also $R_\eta$ would be used before introduced} \SG{ Stimmt, aber ich will es auf einen Blick sichtbar machen }
		\label{tab:PTTparameters}
	\end{table}

We introduce the ratio between the polymer and the solvent viscosity as
\begin{align}
	R_\eta = \frac{\eta_\text{p}}{\eta_\text{s}} = \frac{\eta_0-\eta_\infty}{\eta_\infty}		\label{eq:R_eta}
\end{align}
which will turn out to be an important parameter to determine numerical stability of LB simulations.
\SG{ wo don't actually show that $R_ \eta  $ determines stability? }\RK{we do not. We could do a few simulations with $\alpha_s = 0$ and increasing $R_\eta$ until the sim breaks if you want} \SG{ wie besprochen, bau einfach die Zitate ein, das reicht dann. }\todo{do}
This ratio can similarly be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979} using the zero-shear viscosity $\eta_0$ and the infinite-shear viscosity $\eta_\infty$.
For our alginate solution we obtain $R_\eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological data from various literature sources have been compiled into a single format to determine $R_\eta$.
The \SI{0.49}{\percent} methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.

\begin{table}[H]
	\centering
		\begin{tabular}{l|p{3cm}|p{1.6cm}|p{1.6cm}|p{1.6cm}|p{2cm}}
			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  $>\num{1e4}$ & $<\num{3e-1}$ & $>\num{3e4}$ & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & \num{2e2} & \num{5e-1} & \num{4e2} & Figure 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& \num{5e4} & $<\num{5e1}$ & $>\num{1e3}$ & Fig. 4b \\
			\rowcolor{secondaryBackground}
			O'Connell\cite{oconnell2020} & pluronic F-127 & $>\num{1e6}$ & $<\num{8}$ & $>\num{2e5}$ & Chapter 7 Fig. 13 \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids.
	\SG{ cross-check these values }\RK{Done and formatted}
	}
	\label{tab:bioinks}
\end{table}

@@

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }\RK{dellar is missing}
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }\RK{gupta2015\_2 not in intro} \SG{ dann setz einfach die Zitate so rein, wie du sie fuer passend haeltst. }\todo{do}
Since the early days of LBM, variants for viscoelastic fluids have been developed.
Keeping in mind our target geometries, we focus here on three-dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a basic viscoelastic fluid obeying Jeffery's model \cite{lallemand2003}.
More advanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells.
The work of \cite{wang2020} solved the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM.
The same technique was used by \cite{malaspinas2010} and \cite{ma2020} for the Oldroyd-B and FENE-P models.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \cite{malaspinas2010,ma2020,wang2020,wang2021}, which introduces a small error compared to the analytical solution of the constitutive equation.
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015} \SG{ rephrase and give some more details }\RK{I have not the slightest clue what gupta is doing. I assume optimize is code for increase until stable but I do not know.}.
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
A recent comparison in 2D found that the former in general leads to better stability \cite{dzanic2022_2}.\RK{No in fact, it uses the latter and found to be more accurate than a paper using the former, stability was entirely attributed to the decomposition and artificial diffusivity.} \SG{ die Referenz war falsch, es war der andere Dzanic. Aber da steht eigentlich schon, dass forcing besser sei, oder? }\todo{check/do not mention this}

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\todo{do}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}

@@

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }\RK{Nein}\todo{check}
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. \SG{ comment whether these definitions are identical or different than the definitions above. }\RK{different but can be transformed into each other $\frac{1}{R_\nu}-1=R_\eta$}
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should be 8 instead of 4. \SG{ stimmt das so? }\RK{Ja}
